---
title: "Get started with kelpbio"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get started with kelpbio}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

kelpbio fits Bayesian hierarchical Stan models to estimate kelp biomass and
carbon from field measurements, combining allometric weight, plant size
distribution, density, and wet-to-dry and carbon tissue conversions. This
article will walk through a complete worked example once the modelling
functions are available; for now it serves as a placeholder while the package
API is under development.

## Model checking with the standard Bayesian tooling

A fitted model exposes the `rstantools` generics, so it composes directly with
`bayesplot` and `loo`. The chunks below are illustrative (not run here, as
fitting requires the compiled package).

```{r fit, eval = FALSE}
fit <- kb_fit_weight(kb_data_weight)
```

A posterior-predictive check uses the stored `yrep` via `posterior_predict()`:

```{r ppc, eval = FALSE}
bayesplot::ppc_dens_overlay(
  y = fit$data$weight,
  yrep = posterior_predict(fit)
)
```

Leave-one-out model comparison uses the pointwise `log_lik()`:

```{r loo, eval = FALSE}
loo::loo(log_lik(fit))
```
