---
title: "Get started with kelpbio"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get started with kelpbio}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message = FALSE, warning = FALSE}
library(kelpbio)
```

kelpbio fits Bayesian hierarchical Stan models to estimate kelp biomass from
field measurements. This article walks through the allometric weight model for
*Nereocystis luetkeana*, which relates wet weight to sub-bulb diameter with
random effects for site and site-by-year. Further biomass components (plant size
distribution, density, and wet-to-dry and carbon tissue conversions) are in
development.

Under the hood, models are fitted with [rstan](https://mc-stan.org/rstan/). The
Stan code is pre-compiled when the package is installed, so no separate Stan
installation is required.

## Providing data

The weight model is fitted to a data frame with a numeric `diameter`, a numeric
`weight`, and `site` and `year` grouping factors. A small simulated dataset
is included with the package:

```{r}
data_weight_sim_nereo
```

`kb_check_data_weight_nereo()` confirms the data are in the expected format,
returning the data invisibly if the checks pass and raising an informative error
otherwise:

```{r}
kb_check_data_weight_nereo(data_weight_sim_nereo)
```

## Fitting the model

`kb_fit_weight_nereo()` fits the model. Set `progress = "none"` to silence the
sampler's console output, and `seed` for a reproducible fit. 

```{r, eval = FALSE}
fit <- kb_fit_weight_nereo(
  data_weight_sim_nereo,
  chains = 2,
  niters = 400,
  seed = 42
)
```

```{r, echo = FALSE}
fit <- fit_weight_sim_nereo
```

The short run above keeps the example quick; for a real analysis use the
defaults (`chains = 4`, `niters = 1000`) or more.

## Checking convergence

`glance()` returns a one-row summary of the fit, including the sampler settings
and the convergence diagnostics. `converged` is `TRUE` when the split R-hat is
below its threshold (1.05 by default) and the effective sample size is adequate.

```{r}
glance(fit)
```

## Summarising the fit

`tidy()` returns the model terms with a point estimate and compatibility limits.
The intercept and slopes describe the allometric relationship on the log scale;
the `s`-prefixed terms are the random-effect standard deviations.

```{r}
tidy(fit)
```

## Priors

`kb_priors_weight_nereo()` returns the default priors, which are weakly
informative on the model's centered log scale.

```{r}
kb_priors_weight_nereo()
```

Adjust a prior by replacing an element with `kb_prior_normal()` or
`kb_prior_exponential()` and passing the result to the `priors` argument of
`kb_fit_weight_nereo()`:

```{r, eval = FALSE}
priors <- kb_priors_weight_nereo()
priors$sd_site <- kb_prior_exponential(2)
fit <- kb_fit_weight_nereo(data_weight_sim_nereo, priors = priors, seed = 1)
```

## Predictions

`kb_predict_weight_by()` summarises the fitted weight-diameter curve over a
generated diameter sequence, one curve per group named in `by`. Pass the result
to `kb_plot_predictions()` to plot the curve with its compatibility band:

```{r, fig.width = 7, fig.height = 4.5}
kb_predict_weight_by(fit, by = "site") |>
  kb_plot_predictions()
```

`kb_predict_weight()` instead predicts at the rows of a supplied data frame,
returning a point estimate and compatibility limits for each row:

```{r}
new_data <- data.frame(diameter = c(20, 35, 50, 65, 80), site = "site1")
kb_predict_weight(fit, new_data)
```

## Standard Bayesian tooling

The fitted object exposes the `rstantools` generics, so it composes directly
with `bayesplot` and `loo`. A posterior-predictive check uses the stored `yrep`
via `posterior_predict()`:

```{r, eval = FALSE}
bayesplot::ppc_dens_overlay(
  y = fit$data$weight,
  yrep = posterior_predict(fit)
)
```

Leave-one-out cross-validation uses the pointwise `log_lik()`:

```{r, eval = FALSE}
loo::loo(log_lik(fit))
```
